Post Normalization Repeat of Quality Control Steps
========================================================
  
## Libraries
```{r}
setwd("/big_data/redgar/PAWS/PAWS_final")
library(reshape)
library(ggplot2)
library(RColorBrewer)
library(mclust)
library(IlluminaHumanMethylation450k.db)
library(grid)
library(gridExtra)
```

#### Normalized (BMIQ and probe filtered)
```{r}
load("PAWS_Beta_norm_filtered.RData")
```

#### Cluster function
```{r}
# plot clustering with color function
plotHclustColors <- function(matrix,leafcolor) {
  colnames(matrix) <- leafcolor
  d <- dist(t(matrix))
  hc <- hclust(d, method = "complete") #single, complete, average, ward
  color<-rep(brewer.pal(12,"Paired"),ceiling(length(unique(leafcolor))/12))
  labelColors <- color[sample(1:length(color),length(unique(leafcolor)))]
  colLab <- function(n) {
    if (is.leaf(n)) {
      a <- attributes(n)
      labCol <- labelColors[which(unique(leafcolor) == a$label)]
      attr(n, "nodePar") <- c(a$nodePar, lab.col=labCol)
    }
    n
  }
  clusDendro <- dendrapply(as.dendrogram(hc), colLab)
  plot(clusDendro)
}
```


#### Any meta data
```{r}
metaog<-read.csv("PAWS_sample_info_with_replicates.csv")
metaog$X<-NULL
meta2015<-read.csv("PAWS-GdatasetforKoborlabJan2015-.csv")
meta2015<-meta2015[match(metaog$Psy_ID, meta2015$PAWSG_ID),]
meta2015$Meth_ID<-metaog$Meth_ID
meta2015<-meta2015[,c(1,106, 2:105)]

meta<-meta2015

meta<-meta[which(meta$Meth_ID%in%colnames(PAWS_Beta)),]
meta<-meta[match(colnames(PAWS_Beta), meta$Meth_ID),]

# add genetic cluster to meta from PLINK
Paws_plink_clusters<-read.csv("plink.cluster3.csv", sep="\t")
Paws_plink_four_clusters<-Paws_plink_clusters[,c(2,191)]
colnames(Paws_plink_four_clusters)<-c("Sample_ID","Genetic_cluster")
Paws_plink_four_clusters$Sample_ID<-as.character(Paws_plink_four_clusters$Sample_ID)
Paws_plink_four_clusters$Sample_ID[which(Paws_plink_four_clusters$Sample_ID=="10_9630002082_R10C02")]<-"186"
Paws_plink_four_clusters$Sample_ID[which(Paws_plink_four_clusters$Sample_ID=="10_9630002082_R08C02")]<-"10"
one<-data.frame(Sample_ID=1, Genetic_cluster=0)#add 1 (paws skipped but impute based on dechhieth 4 is 90% 0)
Paws_plink_four_clusters<-rbind(Paws_plink_four_clusters, one)
meta<-merge(meta, Paws_plink_four_clusters, by.x="PAWSG_ID", by.y="Sample_ID")

#add sentrix to meta
sentrix<-read.csv("2014-09-03_pawsSampleSheet_all.csv")
sentrix<-sentrix[,c(1,2,3,6,7)]
rename<-c("3","3F","34","32F","32")
sentrix$Sample_Name<-as.character(sentrix$Sample_Name)
sentrix$Sample_Name[which(sentrix$Sample_Name%in%rename)]
sentrix$Sample_Name[which(sentrix$Sample_Name%in%rename)]<-c("26Fn","34Fn","3n","34n","26n")
meta<-merge(meta, sentrix, by.x="Meth_ID", by.y="Sample_Name")

# want to express sentrix ID as just the row (12 rows)
rows<-c("A","B","C", "D","E","F","G","H")
wells_in_rows<-lapply(rows, function(row) unlist(sapply(c("01","02","03","04","05","06","07","08","09","10","11","12"), function(col) paste(row,col, sep=""))))
wells_in_rows<-as.data.frame(do.call(rbind, wells_in_rows))
meta$plate_row<-sapply(1:nrow(meta), function(well) c(1:12)[apply(wells_in_rows,MARGIN=2,function(x) meta$Sample_Well[well]%in%x)])
meta$plate_row<-sapply(meta$plate_row, function(x) if(length(meta$plate_row[x])==0){NA}else{meta$plate_row[x]})
meta$plate_row<-as.factor(unlist(meta$plate_row))

```

### Confirm Twin IDS and add to meta data
```{r}
### Bring methylumin back in so can check the 65 SNPs on the twins
load("PAWS_methlylumi.RData")
#Sample Changes
Bad_Meth_IDs<-c("100M","40F","109M","120M","95F","92M","134F","136M","50F","167M","26M","31F","27M","27","41_R1","41_R2","34F","60F")
rename<-c("3","3F","34","32F","32")
#filter bad
PAWS.2 <- PAWS.2[,!sampleNames(PAWS.2)%in%Bad_Meth_IDs]#485577 192
#Rename
sampleNames(PAWS.2)[which(sampleNames(PAWS.2)%in%rename)]
sampleNames(PAWS.2)[which(sampleNames(PAWS.2)%in%rename)]<-c("26Fn","34Fn","3n","34n","26n")

PAWS<-PAWS.2
dim(PAWS) # 485577      192 

# Pull out SNP Probes
PAWS_Snps <- PAWS[substring(featureNames(PAWS),1,2) == "rs", ]
PAWS_Snps_betas<-betas(PAWS_Snps)

### load in the twin data
Twin_form_nicki<-read.csv("PAWS-Twin-Data.csv")

meta$PAWSG_ID[which(meta$Meth_ID%in%Bad_Meth_IDs)]
twin<-c(Twin_form_nicki$PAWS.G.ID, Twin_form_nicki$PAWS.G.ID.1)
Pair<-c("TW1","TW2","SB1","TW3","SB2","SB3","TW4","TW5","SB4","TW6","TW7","TW8","TW9",
        "TW1","TW2","SB1","TW3","SB2","SB3","TW4","TW5","SB4","TW6","TW7","TW8","TW9")
fam<-data.frame(Pair, twin)
twin<-fam[which(fam$Pair%in%c("TW1","TW2","TW3","TW4","TW5","TW6","TW7","TW8","TW9")),]
fam_ID<-sapply(1:nrow(meta), function(x) if(meta$PAWSG_ID[x]%in%fam$twin){as.character(fam$Pair)[which(fam$twin==meta$PAWSG_ID[x])]}else{NA})
twin_ID<-sapply(1:nrow(meta), function(x) if(meta$PAWSG_ID[x]%in%twin$twin){as.character(twin$Pair)[which(twin$twin==meta$PAWSG_ID[x])]}else{NA})

meta$Fam_ID<-fam_ID
meta$TWIN_Pair<-twin_ID


PAWS_Snps_betas_clust<-PAWS_Snps_betas[,which(colnames(PAWS_Snps_betas)%in%meta$Meth_ID)]
PAWS_Snps_betas_clust<-PAWS_Snps_betas_clust[,match(meta$Meth_ID, colnames(PAWS_Snps_betas_clust))]
meta[, c(2,113:114)]
plotHclustColors(PAWS_Snps_betas_clust, meta$Fam_ID) #all samples

## Psychip clustering
load("PAWS_snps_Ballelefreq.RData")
SNPs<-PAWS_snps_casted[complete.cases(PAWS_snps_casted),2:ncol(PAWS_snps_casted)] #554034 SNPs
SNPs<-SNPs[,which(colnames(SNPs)%in%meta$PAWSG_ID)]
meta_ordered<-meta[which(meta$PAWSG_ID%in%colnames(SNPs)),]
meta_ordered<-meta_ordered[!duplicated(meta_ordered), ]
meta_ordered<-meta_ordered[match(colnames(SNPs), meta_ordered$PAWSG_ID),]
SNPs_sample<-SNPs[sample(1:nrow(SNPs), 10000),]
SNPs_sample<-as.data.frame(SNPs_sample)
plotHclustColors(SNPs_sample, meta_ordered$Fam_ID) #all samples


save(meta, file="PAWS_meta_sentrix_genetic_clusters.RData")
```









## Load Data
#### Normalized (BMIQ and probe filtered)
```{r}
load("PAWS_Beta_norm_filtered.RData") #192 samples 428407    
load("PAWS_meta_sentrix_genetic_clusters.RData")# with the 2015 data Nicki sent
meta<-meta[which(meta$Meth_ID%in%colnames(PAWS_Beta)),]
meta<-meta[match(colnames(PAWS_Beta), meta$Meth_ID),]
```

# Sample CpGs for plotting and clustering
```{r}
Beta_sample<-PAWS_Beta[sample(1:nrow(PAWS_Beta), 10000),]
```


## Beta distribtuions 
looking good? Take just a random sample of CpGs for plotting (otherwise so slow)
```{r fig.width=10, fig.height=6}
Beta_sample_melted<- melt(Beta_sample)
#remove NAs before plotting (otherwise get many non-inifnite warnings)
Beta_Plot<-Beta_sample_melted[which(!(is.na(Beta_sample_melted$value))),]
#add meta
Beta_Plot<-merge(Beta_Plot,meta, by.x="variable", by.y="Meth_ID")
col<-colorRampPalette(brewer.pal(11,"RdYlGn")[1:8])(4)

ggplot(Beta_Plot, aes(value, group=variable, color=Genetic_cluster))+
  geom_density()+theme_bw()+scale_color_manual(values=col)
```


# Clustering By Any Meta Data variable
```{r fig.width=10, fig.height=3}
# remove rows with NAs
beta_cluster<-Beta_sample[complete.cases(Beta_sample),]

# Plot
plotHclustColors(beta_cluster, meta$Genetic_cluster) #all samples
plotHclustColors(beta_cluster, meta$Fam_ID) #all samples

```





##calculate Tcell proportion (GEO files from Tcells_immunomagnetic_age_matched_PAWS.Rmd) 92 Tcell samples
```{r}

######## Tcell
load("GSE53191.RData")
GSE53191<-GEO[[1]]
GSE53191<-GSE53191[which(rownames(GSE53191)%in%rownames(PAWS_Beta)),]

load("GSE50222.RData")
GSE50222<-GEO[[1]]
GSE50222<-GSE50222[which(rownames(GSE50222)%in%rownames(PAWS_Beta)),]

## filter to the smallest number of CpGs
GSE53191<-GSE53191[which(rownames(GSE53191)%in%rownames(PAWS_Beta)),]
GSE50222<-GSE50222[which(rownames(GSE50222)%in%rownames(PAWS_Beta)),]


meta_tcell<-data.frame(series_id=c(rep("T cells (GSE53191)",12),rep("T cells (GSE50222)",32)),
                 gsm=c(colnames(GSE53191), colnames(GSE50222)),
                 tissue="Tcell",
                 deinc2dep=NA,
                 HGHEDLV2=NA,
                 FADV_6=NA,
                 Genetic_cluster=NA)

GEO_betas_tcell<-cbind(GSE53191, GSE50222)

##### BLOOD
load("GEO_PAWS_aged_matched_whole_blood.RData")

GEO_betas<-GEO_betas[which(rownames(GEO_betas)%in%rownames(PAWS_Beta)),]
PAWS_GEO_meta<-PAWS_GEO_meta[,c(1,2,4)]
PAWS_GEO_meta$series_id<-c(rep("Whole Blood (GSE41169)", 31),rep("Whole Blood (GSE52113)", 6))
PAWS_GEO_meta$deinc2dep<-NA
PAWS_GEO_meta$HGHEDLV2<-NA
PAWS_GEO_meta$FADV_6<-NA
PAWS_GEO_meta$Genetic_cluster<-NA
PAWS_GEO_meta$tissue<-"Whole Blood"


# add tcells to paws
Paws_meta<-meta[,c(1,8,32,20,36,107)]
Paws_meta$series_id<-"Buccal (PAWS)"
Paws_meta$tissue<-"Buccal"
Paws_meta<-Paws_meta[,c(7,1,8,3,4,5,6)]
colnames(Paws_meta)<-c("series_id","gsm","tissue","deinc2dep","HGHEDLV2","FADV_6","Genetic_cluster")


GEO_PAWS_meta<-rbind(Paws_meta, meta_tcell,PAWS_GEO_meta)
GEO_PAWS_beta<-cbind(PAWS_Beta, GEO_betas_tcell, GEO_betas)

save(GEO_PAWS_beta, GEO_PAWS_meta, file="Blood_Tcell_data_with_PAWS.RData")





### PLot whole blood, t cell and PAWS buccals
load("Blood_Tcell_data_with_PAWS.RData")

#PCA
PCA_full_geo<-princomp(na.omit(GEO_PAWS_beta))
Loadings_geo<-as.data.frame(unclass(PCA_full_geo$loadings))
vars <- PCA_full_geo$sdev^2 
vars <- vars/sum(vars) 

## plot
ggplot(Loadings_geo,aes(Comp.1,Comp.2, fill=GEO_PAWS_meta$series_id))+
  geom_point(shape=21, color="black",size=4)+
  theme_bw()+
  scale_fill_manual(values=c("#1d91c0","#9970ab","#c2a5cf","#d9f0a3","#41ab5d"),name="Dataset")+
  xlab("Component 1 (92%)")+ylab("Component 2 (6%)")

# stat
##Check is comp2 as a proxy for blood is correlated with anything
cor.test(Loadings_geo$Comp.2, GEO_PAWS_meta$deinc2dep, method="spearman")
cor.test(Loadings_geo$Comp.2, GEO_PAWS_meta$FADV_6, method="spearman")
cor.test(Loadings_geo$Comp.2, GEO_PAWS_meta$HGHEDLV2, method="spearman")
summary(aov(Loadings_geo$Comp.2~ GEO_PAWS_meta$Genetic_cluster))
```



# HEAT PLOT SCREE
```{r}
### Function of association meta variable with PC (ANOVA)
heat_scree_plot<-function(Loadings, Importance){
  adjust<-1-Importance[1]
  pca_adjusted<-Importance[2:length(Importance)]/adjust
  pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted)))
  
  scree<-ggplot(pca_df[which(pca_df$PC<25),],aes(PC,adjusted_variance))+geom_bar(stat = "identity",color="black",fill="grey")+theme_bw()+
    theme(axis.text = element_text(size =12),
          axis.title = element_text(size =15),
          plot.margin=unit(c(1,1.5,0.2,2.25),"cm"))+ylab("Adjusted Variance")+
    scale_x_continuous(breaks = seq(1,24,1))
  
  #### Heat
  ## correlate meta with PCS
  ## Run anova of each PC on each meta data variable
  
  aov_PC_meta<-lapply(1:ncol(meta_categorical), function(covar) sapply(1:ncol(Loadings), function(PC) summary(aov(Loadings[,PC]~meta_categorical[,covar]))[[1]]$"Pr(>F)"[1]))
  cor_PC_meta<-lapply(1:ncol(meta_continuous), function(covar) sapply(1:ncol(Loadings), function(PC) suppressWarnings(cor.test(Loadings[,PC],as.numeric(meta_continuous[,covar]),alternative = "two.sided", method="spearman", na.action=na.omit)$p.value)))
  names(aov_PC_meta)<-colnames(meta_categorical)
  names(cor_PC_meta)<-colnames(meta_continuous)
  aov_PC_meta<-do.call(rbind, aov_PC_meta)
  cor_PC_meta<-do.call(rbind, cor_PC_meta)
  aov_PC_meta<-rbind(aov_PC_meta, cor_PC_meta)
  aov_PC_meta<-as.data.frame(aov_PC_meta)
  #adjust
  aov_PC_meta_adjust<-aov_PC_meta[,2:ncol(aov_PC_meta)]
  
  #reshape
  avo<-aov_PC_meta_adjust[,1:24]
  avo_heat_num<-apply(avo,2, as.numeric)
  avo_heat<-as.data.frame(avo_heat_num)
  colnames(avo_heat)<-sapply(1:24, function(x) paste("PC",x, sep=""))
  avo_heat$meta<-rownames(avo)
  avo_heat_melt<-melt(avo_heat, id=c("meta"))
  
  # cluster meta data
  meta_var_order<-unique(avo_heat_melt$meta)[rev(ord)]
  avo_heat_melt$meta <- factor(avo_heat_melt$meta, levels = meta_var_order)
  
  # color if sig
  # avo_heat_melt$Pvalue<-sapply(1:nrow(avo_heat_melt), function(x) if(avo_heat_melt$value[x]>=0.9){">=0.9"}else{
  # if(avo_heat_melt$value[x]>=0.5){">=0.5"}else{
  # if(avo_heat_melt$value[x]>=0.1){">=0.1"}else{"<0.1"}}})
  avo_heat_melt$Pvalue<-sapply(1:nrow(avo_heat_melt), function(x) if(avo_heat_melt$value[x]<=0.001){"<=0.001"}else{
    if(avo_heat_melt$value[x]<=0.01){"<=0.01"}else{
      if(avo_heat_melt$value[x]<=0.05){"<=0.05"}else{">0.05"}}})
  
  heat<-ggplot(avo_heat_melt, aes(variable,meta, fill = Pvalue)) +
    geom_tile(color = "black",size=0.5) +
    theme_gray(8)+scale_fill_manual(values=c("#084594","#4292c6","#9ecae1","#deebf7"))+
    theme(axis.text = element_text(size =10, color="black"),
          axis.text.x = element_text(),
          axis.title = element_text(size =15),
          legend.text = element_text(size =14),
          legend.title = element_text(size =12),
          legend.position = c(1, 0), legend.justification = c(1,0),
          plot.margin=unit(c(0,2.25,1,1),"cm"))+
    xlab("Adjusted Principal Component")+ylab(NULL)
  
  grid.arrange(scree, heat, ncol=1, heights = c(2, 4))
}
```


## Run PCA and apply the heat_scree plot function
```{r warning=FALSE, fig.height=9, fig.width=11}
## PCA
PCA_full<-princomp(na.omit(PAWS_Beta))
Loadings<-as.data.frame(unclass(PCA_full$loadings))
vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)



#Specify which covariates are categorical and/or categorical
colnames(meta)
meta$Sentrix_ID<-as.factor(meta$Sentrix_ID)
meta$Genetic_cluster<-as.factor(meta$Genetic_cluster)
meta$DECHIETH<-as.factor(meta$DECHIETH)
meta$Age_Genetic_Collection<-as.factor(meta$Age_Genetic_Collection)
meta$Gender_final<-as.factor(meta$Gender_final)

meta_categorical<-meta[,c("Gender_final","Age_Genetic_Collection","DECHIETH","Genetic_cluster",
                          "Fam_ID","zrank_ord",
                          "Sample_Well","Sample_Plate","Sentrix_ID","Sentrix_Position","plate_row","dateGcollected")]
meta_continuous<-meta[,c("BMI_Percentile","HGHEDLV2","DETOTIN2","COMPSES2","deinc2dep","FLadder","FADV_6")]
c(colnames(meta_categorical),colnames(meta_continuous))
ord <- c(1,2,3,4,5,6,12,13,14,15,16,17,18,19,7,8,9,10,11)


#Apply function on PCA results, pulls in the meta data and beta values from above
heat_scree_plot(Loadings, Importance)

# COMBAT: Sentrix_ID, Sentrix_Position
# LME covairates: Age_Genetic_colletion, Fam_ID, Genetic Cluster, 
```




## meta data correlation for covariate selection in lm
```{r}
meta_categorical<-meta[,c("Gender_final","Age_Genetic_Collection","DECHIETH","Genetic_cluster",
                          "zrank_ord","minor_child",
                          "Sample_Well","Sample_Plate","Sentrix_ID","Sentrix_Position","plate_row","dateGcollected")]
meta_continuous<-meta[,c("BMI_Percentile","HGHEDLV2","DETOTIN2","COMPSES2","deinc2dep","FLadder","FADV_6")]


meta_cor<-cbind(meta_categorical,meta_continuous)

Meta_correlation<-lapply(1:ncol(meta_cor), function(x) as.numeric(meta_cor[,x]))
Meta_correlation<-as.data.frame(do.call(cbind, Meta_correlation))
colnames(Meta_correlation)<-colnames(meta_cor)

correlation<-as.data.frame(cor(Meta_correlation, use="complete.obs", method="spearman") )

# choose number of clusters
wss <- (nrow((correlation))-1)*sum(apply((correlation),2,var))
for (i in 2:10) wss[i] <- sum(kmeans((correlation), 
                                     centers=i)$withinss)
plot(1:10, wss, type="b", xlab="Number of Clusters",
     ylab="Within groups sum of squares")

k <- kmeans(correlation,10)

# Append id and cluster
correlationc <- cbind(correlation, id=seq(nrow(correlation)), cluster=k$cluster)
correlationc$idsort <- correlationc$id[order(correlationc$cluster)]
correlationc$idsort <- order(correlationc$idsort)
correlationc$cluster <- NULL

# use reshape2::melt to create data.frame in long format
correlationc$id<-rownames(correlationc)
correlationm <- melt(correlationc, id.vars=c("id","idsort"))
library(scales)
cols <- brewer_pal(pal = "RdBu")(5)

ggplot(correlationm, aes(id, variable, fill = value)) +
  geom_tile(color = "black",size=0.5) +
  theme_gray(8)+theme(axis.text = element_text(size =10, color="black"),
          axis.title =  element_blank(),
          axis.text.x = element_text(angle = 90, hjust = 1, , vjust=0.5))+
  scale_fill_gradientn(colours = cols, 
                         values = rescale(c(-1, 0,1)),
                         guide = "colorbar", limits=c(-1, 0.99))
```


# Confounding of minority status and genetic cluster for inclusion in models of vairables
# should maybe be done after combat
```{r}
load("PAWS_Beta_norm_filtered.RData") #192 samples 428407    
load("PAWS_meta_sentrix_genetic_clusters.RData")# with the 2015 data Nicki sent
meta<-meta[which(meta$Meth_ID%in%colnames(PAWS_Beta)),]
meta<-meta[match(colnames(PAWS_Beta), meta$Meth_ID),]

# Minority status
meta_minor<-meta[which(!(is.na(meta$minor_child))),]
meta_minor$minor_child<-as.factor(meta_minor$minor_child)
meta_minor_melt<-melt(meta_minor[,c("Meth_ID","minor_child","HGHEDLV2","COMPSES2","DETOTIN2",
                                    "deinc2dep","FLadder","FADV_6","CF1rungE",
                                    "CF1rungC","zrank_ord","SPNILadb")], id=c("Meth_ID","minor_child"))
meta_minor_melt<-meta_minor_melt[which(meta_minor_melt$value<600),]

ggplot(meta_minor_melt, aes(minor_child,value))+
  geom_violin(aes(fill=minor_child))+
  geom_boxplot(fill="grey",width=0.1, outlier.shape=19, outlier.size=1)+
  facet_wrap(~variable, scales="free_y", ncol=5)+
  theme_bw()+scale_fill_manual(values=c("#ef8a62","#67a9cf"), name="Minority Status")+
  xlab("Child Ethnic Minority Status")
    
interest<-c("Meth_ID","minor_child","HGHEDLV2","COMPSES2","DETOTIN2",
                                    "deinc2dep","FLadder","FADV_6","CF1rungE",
                                    "CF1rungC","zrank_ord","SPNILadb") 
pval<-data.frame(variable=interest[3:length(interest)],
                 pval=sapply(3:length(interest),function(x){
         round(t.test(meta_minor[which(meta_minor$minor_child==0),interest[x]],
                meta_minor[which(meta_minor$minor_child==1),interest[x]])$p.value,4)}))



#Genetic Cluster confounds
meta_GC<-meta[which(!(is.na(meta$Genetic_cluster))),]
meta_GC_melt<-melt(meta_GC[,c("Meth_ID","Genetic_cluster","HGHEDLV2","COMPSES2","DETOTIN2",
                                    "deinc2dep","FLadder","FADV_6","CF1rungE",
                                    "CF1rungC","zrank_ord","SPNILadb")], id=c("Meth_ID","Genetic_cluster"))
col<-colorRampPalette(brewer.pal(11,"RdYlGn")[1:8])(4)
meta_GC_melt<-meta_GC_melt[complete.cases(meta_GC_melt),]
meta_GC_melt$Genetic_cluster<-as.factor(meta_GC_melt$Genetic_cluster)
meta_GC_melt<-meta_GC_melt[which(meta_GC_melt$value<600),]

ggplot(meta_GC_melt, aes(Genetic_cluster,value))+
      geom_violin(aes(fill=Genetic_cluster))+
      geom_boxplot(fill="grey",width=0.1, outlier.shape=19, outlier.size=1)+
  facet_wrap(~variable, scales="free_y", ncol=5)+scale_fill_manual(values=col, name="Genetic Cluster")+
      theme_bw()+
      xlab("Genetic Cluster")
interest<-c("Meth_ID","Genetic_cluster","HGHEDLV2","COMPSES2","DETOTIN2",
                                    "deinc2dep","FLadder","FADV_6","CF1rungE",
                                    "CF1rungC","zrank_ord","SPNILadb") 

pval<-data.frame(variable=interest[3:length(interest)],
                 pval=sapply(3:length(interest),function(x){
         round(summary(aov(meta_GC[,interest[x]]~meta_GC[,"Genetic_cluster"]))[[1]][["Pr(>F)"]][1],4)}))



## Age
meta_age<-meta[which(!(is.na(meta$Age_Genetic_Collection))),]
meta_age_melt<-melt(meta_GC[,c("Meth_ID","Age_Genetic_Collection","HGHEDLV2","COMPSES2","DETOTIN2",
                                    "deinc2dep","FLadder","FADV_6","CF1rungE",
                                    "CF1rungC","zrank_ord","SPNILadb")], id=c("Meth_ID","Age_Genetic_Collection"))
meta_age_melt<-meta_age_melt[complete.cases(meta_age_melt),]
meta_age_melt<-meta_age_melt[which(meta_age_melt$value<600),]
meta_age_melt$Age_Genetic_Collection<-as.numeric(as.character(meta_age_melt$Age_Genetic_Collection))

ggplot(meta_age_melt, aes(Age_Genetic_Collection,value))+
  geom_point(color="grey",shape=19)+
  facet_wrap(~variable, scales="free_y", ncol=5)+
  theme_bw()+
  xlab("Age Genetic Collection")+  stat_smooth(method="lm")

interest<-c("Meth_ID","Age_Genetic_Collection","HGHEDLV2","COMPSES2","DETOTIN2",
                                    "deinc2dep","FLadder","FADV_6","CF1rungE",
                                    "CF1rungC","zrank_ord","SPNILadb") 

pval_lm<-data.frame(variable=interest[c(5,7,9,12)],
                 pval=sapply(c(5,7,9,12),function(x){
         round(coef(summary(lm(meta_age[,interest[x]]~meta_age[,"Age_Genetic_Collection"])))[2,"Pr(>|t|)"],4)}))
pval_aov<-data.frame(variable=interest[c(3,4,6,8,10,11)],
                     pval=sapply(c(3,4,6,8,10,11),function(x){
         round(summary(aov(meta_age[,interest[x]]~meta_age[,"Age_Genetic_Collection"]))[[1]][["Pr(>F)"]][1],4)}))



## gender
meta_gender<-meta[which(!(is.na(meta$Gender_final))),]
meta_gender_melt<-melt(meta_GC[,c("Meth_ID","Gender_final","HGHEDLV2","COMPSES2","DETOTIN2",
                                    "deinc2dep","FLadder","FADV_6","CF1rungE",
                                    "CF1rungC","zrank_ord","SPNILadb")], id=c("Meth_ID","Gender_final"))
meta_gender_melt<-meta_gender_melt[complete.cases(meta_gender_melt),]
meta_gender_melt<-meta_gender_melt[which(meta_gender_melt$value<600),]
meta_gender_melt$Gender_final<-as.factor(meta_gender_melt$Gender_final)

ggplot(meta_gender_melt, aes(Gender_final,value))+
      geom_violin(aes(fill=Gender_final))+
      geom_boxplot(fill="grey",width=0.1, outlier.shape=19, outlier.size=1)+
  facet_wrap(~variable, scales="free_y", ncol=5)+
      theme_bw()+scale_fill_manual(values=c("#7fcdbb","#9970ab"), name="Gender")+
      xlab("Gender")

interest<-c("Meth_ID","Gender_final","HGHEDLV2","COMPSES2","DETOTIN2",
                                    "deinc2dep","FLadder","FADV_6","CF1rungE",
                                    "CF1rungC","zrank_ord","SPNILadb") 

pval<-data.frame(variable=interest[3:length(interest)],
                 pval=sapply(3:length(interest),function(x){
         round(t.test(meta_gender[which(meta_gender$Gender_final==0),interest[x]],
                meta_gender[which(meta_gender$Gender_final==1),interest[x]])$p.value,4)}))

```





#Combat
```{r}
library(sva)
library(pamr)
library(limma)


# COMBAT: Sentrix_ID, plate_row, Sentrix_Position (cant correct dateGcollected, many with just one per date)
# LME covairates: Age_Genetic_colletion, Fam_ID, Genetic Cluster
# Protect: HGHEDLV2+DETOTIN2+COMPSES2+deinc2dep+FLadder+FADV_6+zrank_ord

# EL_CG data
pheno = meta
pheno$Sentrix_ID<-as.factor(pheno$Sentrix_ID)

Mval<-function(beta) log2(beta/(1-beta))
edata = apply(PAWS_Beta, 1, Mval) # need mvalues for combat
edata = as.data.frame(edata)
edata = t(edata)

#Combat
#Combat can't deal with NAs in meta data Impute pheno
f=function(x){
   x<-as.numeric(as.character(x)) #first convert each column into numeric if it is from factor
   x[is.na(x)] =median(x, na.rm=TRUE) #convert the item with NA to median value from the column
   x #display the column
}
pheno$HGHEDLV2<-f(pheno$HGHEDLV2)
pheno$COMPSES2<-f(pheno$COMPSES2)
pheno$deinc2dep<-f(pheno$deinc2dep)
pheno$FLadder<-f(pheno$FLadder)
pheno$FADV_6<-f(pheno$FADV_6)
pheno$zrank_ord<-f(pheno$zrank_ord)
pheno$DETOTIN2<-f(pheno$DETOTIN2)

#this model matrix does work                      nCovs=(3,4,5,6)
mod = model.matrix(~as.factor(HGHEDLV2)+DETOTIN2+COMPSES2+deinc2dep+FLadder+FADV_6+zrank_ord, data=pheno) 
numCovs<-c(7,8,9,10,11,12)

## filter out CpGs which have NAs batched by sentrix Id
sentrix_na<-sapply(1:nrow(edata), function(cpg) max(tapply(edata[cpg,], pheno$Sentrix_ID, function(x) sum(is.na(x)))))
edata_filtered<-edata[which(sentrix_na<3),]# 426893 (1514 CpGs had to be removed because NAs batched by sentrix ID)

#Sentrix_ID
batch = pheno$Sentrix_ID
combat_PAWS_Beta = ComBat(dat=edata_filtered, batch=batch, mod=mod, numCovs=numCovs, par.prior=TRUE)



#plate_row only need if SENtrix ID doesnt correct it too
batch = pheno$Sentrix_Position
combat_PAWS_Beta = ComBat(dat=combat_PAWS_Beta, batch=batch, mod=mod, numCovs=numCovs, par.prior=TRUE)
combat_PAWS_Mval<-combat_PAWS_Beta

#Back to betas
betas<-function(M) 2^M/((2^M)+1)
combat_PAWS_Beta = apply(combat_PAWS_Beta, 1, betas) # need mvalues for combat
combat_PAWS_Beta = as.data.frame(combat_PAWS_Beta)
combat_PAWS_Beta = t(combat_PAWS_Beta)
combat_PAWS_Beta<-combat_PAWS_Beta

save(combat_PAWS_Beta, combat_PAWS_Mval, file="PAWS_Combatted.RData")
```





# Compare Pre and post normalization and combat REPLICATES
```{r}
load("PAWS_Combatted.RData") # Includes replicates
load("PAWS_methlylumi.RData")
load("PAWS_sampFiltered_prbFilter_BMIQ.RData")

beta_raw<-betas(PAWS.2)
rename<-c("3","3F","34","32F","32")
colnames(beta_raw)[which(colnames(beta_raw)%in%rename)]<-c("26Fn","34Fn","3n","34n","26n")
beta_bmiq<-betas(PAWS)
beta_Combat<-combat_PAWS_Beta

Replicates<-c("33F","33","180F","180","186F","186","26Fn","26n","34Fn","34n","10F","10","104M","104","116M","116","117F","117rF","121F","121","72F","72","78rF","78F","20rM","20r2M","20M")# 13 replicated samples one sample has 3 replicates total (20M)

original_rep<-beta_raw[,which(colnames(beta_raw)%in%Replicates)]
bmiq_rep<-beta_bmiq[,which(colnames(beta_bmiq)%in%Replicates)]
combat_rep<-beta_Combat[,which(colnames(beta_Combat)%in%Replicates)]



replicated_individuals<-c("33","180","186","26","34","10","104", "116","117","121","72","78","20")


###Root mean squared error
library(hydroGOF)
Data_rmse<-function(df){
  rep_cor<-do.call(rbind,lapply(1:12, function(x) {cols<-grep(replicated_individuals[x], colnames(df))
                                                   data.frame(Comparison=replicated_individuals[x],
                                                              Correlation=rmse(df[,cols[1]],
                                                                              df[,cols[2]],use="complete.obs"))}))
  rep_cor_20<- data.frame(Comparison=replicated_individuals[13],
                          Correlation=c(rmse(df[,5],df[,10]),
                                        rmse(df[,5],df[,17]),
                                        rmse(df[,10],df[,17])))
  rep_cor<-rbind(rep_cor, rep_cor_20)}

original_rep_cor<-Data_rmse(original_rep)
bmiq_rep_cor<-Data_rmse(bmiq_rep)
combat_rep_cor<-Data_rmse(combat_rep)

original_rep_cor$Data<-"Original"
bmiq_rep_cor$Data<-"BMIQ Norm"
combat_rep_cor$Data<-"BMIQ Norm Combat"


rep_cor<-rbind(original_rep_cor,bmiq_rep_cor,combat_rep_cor)
rep_cor$Data <- factor(rep_cor$Data, levels=c("Original","BMIQ Norm", "BMIQ Norm Combat"))
rep_cor$Comparison<-as.character(rep_cor$Comparison)
rep_cor$Comparison[43:45]<-c("20","20_1","20_2")
rep_cor$Comparison[28:30]<-c("20","20_1","20_2")
rep_cor$Comparison[13:15]<-c("20","20_1","20_2")

save(rep_cor, file="replicate_RMSE.RData")


ggplot(rep_cor, aes(Data, Correlation, group=Comparison, color=Comparison))+geom_line(size=2)+theme_bw()+ylab("Replicate Root Mean Squared Error in Beta Value")


Replicates<-c("33F","180F","186F","26Fn","34Fn","10F","104M", "116M","117F","121F","72F","78rF","20rM","20r2M")# use later will mean 176 unique samples
combat_PAWS_Beta_norep<-combat_PAWS_Beta[,which(!(colnames(combat_PAWS_Beta)%in%Replicates))]#426893    178
PAWS_Beta<-combat_PAWS_Beta_norep
save(PAWS_Beta, file="combat_PAWS_Beta_norep.RData")
```



# PCA after COMBAT
```{r}
## load EL_CG data, meta and Hannum
load("combat_PAWS_Beta_norep.RData")
load("PAWS_meta_sentrix_genetic_clusters.RData")

meta<-meta[which(meta$Meth_ID%in%colnames(PAWS_Beta)),]
meta<-meta[match(colnames(PAWS_Beta), meta$Meth_ID),]

PCA_full_combat<-princomp(na.omit(PAWS_Beta))
Loadings_combat<-as.data.frame(unclass(PCA_full_combat$loadings))
vars_combat <- PCA_full_combat$sdev^2
Importance_combat<-vars_combat/sum(vars_combat)

meta$Sentrix_ID<-as.factor(meta$Sentrix_ID)
meta$Genetic_cluster<-as.factor(meta$Genetic_cluster)
meta$DECHIETH<-as.factor(meta$DECHIETH)
meta$Age_Genetic_Collection<-as.factor(meta$Age_Genetic_Collection)
meta$Gender_final<-as.factor(meta$Gender_final)

meta_categorical<-meta[,c("Gender_final","Age_Genetic_Collection","DECHIETH","Genetic_cluster",
                          "Fam_ID","zrank_ord",
                          "Sample_Well","Sample_Plate","Sentrix_ID","Sentrix_Position","plate_row","dateGcollected")]
meta_continuous<-meta[,c("BMI_Percentile","HGHEDLV2","DETOTIN2","COMPSES2","deinc2dep","FLadder","FADV_6")]
c(colnames(meta_categorical),colnames(meta_continuous))
ord <- c(1,2,3,4,5,6,12,13,14,15,16,17,18,19,7,8,9,10,11)

heat_scree_plot(Loadings_combat, Importance_combat)
```

